Kindly Bent to Free Us

### 2019-07-09 kindly/soa.tex
SOA²

Manuel Fahndrich and Robert DeLine. Adoption and focus: Practical
linear types for imperative programming. In PLDI, pages 13–24, 2002.

\subsubsection{Functional linear of affine type systems}

@@

Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
to retrofit Haskell with linear types.
Linear Haskell introduces a new notion
of ``linear arrows'', noted $\multimap$ similar to linear logic,
which are functions who use their argument exactly once.
By contrast, in \lang linearity is only decided
by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has several disadvantages\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}
\item Their system does not admit subtyping for arrows and leverages
  $\eta$-expansion to allow passing unrestricted functions in linear
  contexts. This is acceptable in a non-strict language such as
  Haskell but can not be used in a strict setting.
\item
  While the GHC proposal implements some type inference,
  their algorithm is not formalized and not proved complete.
\item
  Linear Haskell promotes a CPS-style such as 
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure that files are only used linearly.
  \lang follows System F\degree, Quill and Alms and
  adopts a direct style, where types themselves are
  described as affine/linear.
  We believe this leads to a more natural programming style
  and ease modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need consider the module
  implementation to ensure that some linear
  resource is properly handled.
\item
  Linear Haskell introduces a notion of linear monad to express
  imperative code conveniently. Again, this solution is suitable in Haskell,
  but less appropriate in other contexts. Our borrow system allow
  to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

\paragraph{Vault\citep{DBLP:conf/pldi/DeLineF01,DBLP:conf/pldi/FahndrichD02}}

\subsubsection{Borrows}

Rust, Non Lexical lifetimes

\subsubsection{Object systems}
\TODO{}

@@

\subsubsection{Uniqueness}

\subsubsection{Others}

\paragraph{Mezzo}

\subsection{Qualified Types}

HM(X)~\citep{DBLP:journals/tapos/OderskySW99} with extensions to
bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
and GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.

\subsection{Subtyping}

Flowcaml~\citep{DBLP:conf/popl/PottierS02},
\citet{DBLP:conf/sas/TrifonovS96}.

>>>>

\subsubsection{Functional languages}

@@

Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
to retrofit Haskell with linear types.
Linear Haskell introduces a new notion
of ``linear arrows'', noted $\multimap$ similar to linear logic,
which are functions who use their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has several disadvantages\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}
\item Their system does not admit subtyping for arrows and leverages
  $\eta$-expansion to allow passing unrestricted functions in linear
  contexts. This is acceptable in a non-strict language such as
  Haskell but can not be used in a strict setting.
\item
  While the GHC proposal implements some type inference,
  their algorithm is not formalized and is not complete.
\item
  Linear Haskell promotes a CPS style with functions such as 
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure that files are only used linearly.
  \lang follows System F\degree, Quill and Alms and
  adopts a direct style, where types themselves are
  described as affine or linear.
  We believe this leads to a more natural programming style
  and ease modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need consider the module
  implementation to ensure that some linear
  resource is properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
with an extremely rich capability system which allows it to encode numerous
properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
The main contribution of Mezzo is to push the boundary of what is possible
to do with a type system. This makes it far more expressive than \lang, but
also quite harder to use, and precludes any form of complete type
inference.

\subsubsection{Other systems}

Vault~\citep{DBLP:conf/pldi/DeLineF01}
is a low-level language which uses typestate along with \emph{keys}
to track ownership. It also introduces a novel paradigm
called \emph{adoption and focus} \citep{DBLP:conf/pldi/FahndrichD02}
which allow restricting the scope of linear objects
(adoption) or temporarily treating them as non-linear (focus).
These mechanisms serve a similar purpose as borrows, but are tracked
much more precisely thanks to the use of typestate and pre/post-conditions,
which makes it very expressive (at the cost of some verbosity).
However, Vault is mostly geared towards traditional imperative programming:
functions can't capture linear objects and it does not support
type inference.

Rust~\citep{rust} is the first
mainstream language to popularize the idea of borrowing and ownership
for safe low-level programming.
\lang heavily draws inspiration from Rust's borrowing system to apply
it in a functional setting with type inference, a GC, and
an ML module system.
Rust's lifetime system is more explicit and more expressive than \lang,
however Rust does not provide type inference
and only provides \emph{partial} lifetime inference.
Recently, \citet{DBLP:journals/corr/abs-1903-00982} presented Oxide,
a core language for Rust which formalizes its ownership discipline
including non-lexical lifetimes, which we look forward to use to improve
our borrowing system.

@@

\subsection{Type-system features}

\lang strongly relies on qualified types~\cite{DBLP:journals/scp/Jones94}
to introduce the necessary kind equalities in our system. In
particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
allow to use qualified type in an ML-like language with complete
type inference.
HM(X) has been shown to be compatible with subtyping,
bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.
This makes us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml.

The subtyping discipline we develop in \lang is very similar
to structural subtyping, where the only subtyping (or here, subkinding)
is at the leafs.
Such discipline is known to be friendly to inference and has be used in many
contexts, including OCaml itself, and has been combined
with constraints~\citet{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
extends OCaml with security levels forming a lattice and supports type inference.
In \lang, we repurpose these ideas to check for linearity and affinity.
\citet{DBLP:conf/aplas/Simonet03} also presents various algorithmic
notions to solve lattice-based constraints, which we partially use
in our constraint solving algorithm. Our main contribution
is the design of a kind language specialized for checking linearity
and sufficient simple to make
all simplification rules complete, which allows to keep our type signature simple.
